---
title: "Research Analysis Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

```{r}
#| label: data load
load(here("data","data_constructed","analytical_data.RData"))
```

```{r}
#| label: tbl-multi-models
#| tbl-cap: Multinomial Models
#| results: asis
model1 <- multinom(race_response~HomelandP,
                  data=subset(DATA, race_ancestry=="Indigenous/White"))
model2 <- multinom(race_response~HomelandP+EducA,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
model3 <- multinom(race_response~HomelandP+EducA+Gender,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
model4 <- multinom(race_response~HomelandP+EducA+Gender+Hispanic,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
model5 <- multinom(race_response~HomelandP+EducA+Gender+Hispanic+MetroC,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
knitreg(list(model1,model2,model3,model4,model5),
          digits=3)
```
Looking at model 1 can see that living on Homeland leads to increased odds of identifying as Indigenous alone, and this remains consistent through various iterations of models until the 
```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -1)) 
```

In @fig-race-ancestry 72% of respondents who list Indigenous and Asian ancestry did not self-identify as Indigenous or Asian in regards to their racial identity. \^[Does this challenge previous research that suggests that those with part Asian ancestry tend to favor a multiracial identity @gullickson_choosing_2011. This can help lead to some future analysis and inform the inclusion of more literature on multiracial identity amongst mixed-race Asians]
