---
title: "Navigating Ancestry and Racial Classification Amongst Multiracial Indigenous American"
shorttitle: "Navigating Ancestry and Racial Classification"
abstract: Current research on race and multiracial individuals recognizes that Indigenous racial identity is fluid and often contested. Using a settler-colonial theoretical framework and the recommendations of QuantCrit literature, this research project expands understanding of multi-racial Indigenous identity and demonstrates how ancestry influences racial “decision-making” for Indigenous Americans. By leveraging data from the American Community Survey (ACS) from 2010 to 2020, the relationship between an individuals reported ancestry and their self-identified racial classification is investigated by estimating multinomial logistic regression models. The results indicate the relationship between various predictorss, and a persons likelihood to identify as Indigenous alone, multi-racial, or to distance themself from their Indigenous ancestry and identity in favor of their other racial identity. When these findings are evaluated within the context of settler-colonialism, many previously confounding findings can be linked to the social reality that respondents navigate as Indigenous Americans.
keywords: [Indigenous identity, multiracial studies, ancestry, racial decision-making, racial classification, settler-colonialism, QuantCrit, American Community Survey]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/apa-6th-edition.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData"))
```

# Introduction

Need to fix intro still but here is a start

One group of scholars has attempted to better understand Indigenous identity by determining which factors are associated with strong Indigenous identity, how this identity is contested and understood at an interpersonal and community level, and how the Indigenous population has grown over time (Liebler, 2010; Lucero, 2014; McKay, 2021). Indigenous identity becomes even more complicated when considering multiracial Indigenous Americans, which has led to additional research in this area, although limited. Articles on Indigenous identity generally cover the racial decision-making process for multiracial Indigenous people in the United States, since many Indigenous people have mixed racial ancestries. This social reality requires Indigenous American respondents to make conscious decisions when listing their race. In the case of multiracial Indigenous people their known racial ancestry either conflicts with or affirms their answer, and they must reconcile with that decision. Understanding this decision-making process is crucial for gaining insights into contemporary shafts in racial boundary formation and the individual considerations that shape racial identification for Indigenous Americans.

However, in 2020 Dwanna L. McKay, Kirsten Vinyeta, and Kari M. Norgaard published an article titled “Theorizing Race and Settler Colonialism within U.S. Sociology” which calls for a critical approach to how race is studied within Sociology, particularly how the focus on assimilation fails to confront its violent and often deadly nature of assimilation for Indigenous Americans under settler colonialism. Their article established that as researchers and academics, we are complicit in the erasure of Indigenous people, and the continued reification of harmful narratives can only be stopped by engaging with settler colonialism theory when conducting our research.

These considerations led to the theoretical engagement explored later in this paper to prepare for a critical and nuanced approach to quantitative research has often been critiqued for its inability to speak to individual experiences, and for those who study race, there have been critiques of using race as a static category that does not reflect the social reality of race as a social construct, which parallels the arguments mentioned above about settler colonialism within the sociology of race and ethnicity. This dynamic nature of race means that it can and does change, and is influenced by various social factors and processes at both the micro and macro level which is why a critical approach to the interpretation and contextualization of results from quantitative research is imperative to furthering understanding without sacrificing reflexivity (Zuberi & Bonilla-Silva, 2008).

# Theoretical Framework

In 2018 the journal of Race Ethnicity and Education published a special issue titled "QuantCrit: Rectifying Quantitative Methods Through Critical Race Theory", where various scholars critically engaged with quantitative research and methods, using CRT to determine whether quantitative research is capable of being used to further a critical race agenda in education research and beyond (Garcia et al., 2018). This call for more critically engaged quantitative research is at the basis of this article, meaning that the results will have been interpreted within the context of CRT as well as settler-colonialism due to its focus on Indigenous people. This approach is not new, however, and has already been established in Du Bosian sociology as a

Middle section major work in progress

Furthermore, the interpretation of results through the lens of CRT, QuantCrit, or settler-colonialism may not be able to account for every finding however engaging with the social reality Indigenous Americans navigate only serves to expand and explain current understandings regarding Indigenous identity, particularly multi-racial Indigenous identity

# Methods and data

## Sample Selection and Data

The data used for this project was compiled from the American Community Survey data for the years 2010 - 2020. I seek to analyze the race reporting of individuals who indicate Indigenous ancestry in addition to White, Black, or Asian ancestry. For this study, the term Indigenous is used in place of the American Indian and Alaskan Native (AIAN) terms as it is generally more accepted and has grown in popularity in current literature when discussing identity. The sample did not include individuals who listed their race as Latino or Pacific Islander.Part of the data organization of this project required the identification of a multiracial population, specifically an Indigenous multiracial population. To accomplish this goal the Goldstein and Morning method was adopted, which relies on looking at adults' self-reported ancestry, which in turn does not rely on the question of racial identity and instead allows for the construction of a multiracial population based on respondents' self-reported racial ancestry. In regards to the sample, this means looking at respondents who listed Indigenous ancestry in addition to a racially distinct racial ancestry other than Indigenous.Once this group was identified further case selection was performed as these results were compared to their race response. When looking at their racial ancestry and race response, respondents were divided into 4 distinct categories:1) Indigenous alone, which contained respondents that listed their race solely as Indigenous 2) Consistent which included respondents that acknowledged their racial ancestry in full by selecting a racial option that reflected that ancestry 3) Non-Indigenous Race alone, where respondents did acknowledge their Indigenous ancestry when selecting their race response and 4) Inconsistent, where respondents race response did not reflect their racial ancestry in any capacity

Outcome variable: race response Predictor variables: metro, homeland, education above high school, gender and hispanicity These ancestry responses were compared to what their race response and divided into the following groups: Indigenous alone, Multiracial, and Other Race alone where Indigenous alone consisted of those respondents who identified a mixed-ancestry of Indigenous plus one other racial ancestry yet listed their race response as Indigenous alone. Multiracial was those respondents who identified their mixed--Indigenous ancestry and listed a multiracial race response. Other race alone consisted of those respondents with a mixed-Indigenous racial ancestry but listed their racial response as the other race (ie. ancestry listed as Indigenous and Asian but list race as Asian only).

## Methods

The analysis was conducted through creation of several multinomial logit models, which predict the likelihood of a respondent to identify as either Indigenous alone, Multiracial, or other race alone. Three sets of models were created, each set had a different reference category: "Indigenous/White", "Indigenous/Black", and "Indigenous/Asian"

```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -2))
```

```{r}
```

# Results

# Conclusion
