---
title: "Navigating Ancestry and Racial Classification Amongst Multiracial Indigenous American"
shorttitle: "Navigating Ancestry and Racial Classification"
abstract: Current research on race and multiracial individuals recognizes that Indigenous racial identity is fluid and often contested. Using a settler-colonial theoretical framework and the recommendations of QuantCrit literature, this research project expands understanding of multi-racial Indigenous identity and demonstrates how ancestry influences racial “decision-making” for Indigenous Americans <!-- COMMENT: The ancestry question is your lever into being able to do the work, but is it being used to demonstrate anything? I also think its unclear from the sentence that you are specifically looking at the decision making of individuals of multiracial indigenous ancestry, rather than indigenous individuals more broadly. You want to be real clear to the reader what you are exactly doing in the abstract. --> . By leveraging data from the American Community Survey (ACS) from 2010 to 2020, the relationship between an individuals reported ancestry and their self-identified racial classification is investigated by estimating multinomial logistic regression models. The results indicate the relationship between various predictorss, and a persons likelihood to identify as Indigenous alone, multi-racial, or to distance themself from their Indigenous ancestry and identity in favor of their other racial identity. When these findings are evaluated within the context of settler-colonialism, many previously confounding findings can be linked to the social reality that respondents navigate as Indigenous Americans.
keywords: [Indigenous identity, multiracial studies, ancestry, racial decision-making, racial classification, settler-colonialism, QuantCrit, American Community Survey]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/apa-6th-edition.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData"))
```

## Introduction

Who is Indigenous and who is not? Where do multiracial Indigenous Americans fit in that equation? One group of scholars has asked similar questions in an attempt to better understand Indigenous identity by determining which factors are associated with strong Indigenous identity, how this identity is contested and understood at an interpersonal and community level, and how the Indigenous population has grown over time (Liebler, 2010; Lucero, 2014; McKay, 2021). Indigenous identity becomes even more complicated when considering multiracial Indigenous Americans, which has led to additional research in this area, although limited. Articles <!-- COMMENT: I think "Research" would be better here than "Articles" --> on Indigenous identity generally cover the racial decision-making process for multiracial Indigenous people in the United States, since many Indigenous people have mixed racial ancestries <!-- COMMENT: some citations would be good here. You are touching on a big literature! -->. This social reality requires Indigenous American respondents to make conscious decisions when listing their race. In the case of multiracial Indigenous people their known racial ancestry either conflicts with or affirms their answer, and they must reconcile with that decision. <!-- COMMENT: I don't full understand what you are saying here. It seems like you are presuming there is a struggle but we don't know that this is the case. More importantly, I think the reader will be confused by the way you are using ancestry here because you haven't established the technical use of that term methodologically. -->  Understanding this decision-making process is crucial for gaining insights into contemporary shifts in racial boundary formation and the individual considerations that shape racial identification for Indigenous Americans <!-- and sometimes non-indigenous Americans, in the sense that far more people identify as of indigenous ancestry than of indigenous race! -->. 

Dwanna L. McKay, Kirsten Vinyeta, and Kari M. Norgaard published an article titled “Theorizing Race and Settler Colonialism within U.S. Sociology” which calls for a critical approach to how race is studied within Sociology, particularly how the focus on assimilation fails to confront its violent and often deadly nature for Indigenous Americans under settler colonialism (2020). Their article established that as researchers and academics, we are complicit in the erasure of Indigenous people, and the continued reification of harmful narratives can only be stopped by engaging with settler colonialism theory when conducting our research. <!-- COMMENT: in the context of what comes before, this paragraph just seems kind out on its own. Its not clear how it relates to your topic or what you are doing to specifically address it in your research approach. This is to some extent true in the paragraph below as well. I know you want to get into it more later, but what you are trying to do here in these two paragraphs could probably be much briefer and also tie back into what you are doing more specifically. I think you are taking a constructivist approach to race in the method here and that is worth saying at the outset. --> 

These considerations led to the theoretical engagement explored later in this paper to prepare for a critical and nuanced approach to quantitative research that has often been critiqued for its inability to speak to individual experiences <!-- COMMENT: Why should social science work ever speak to individual experience? -->. For those who study race, there have been criticisms of using race as a static category that does not reflect the social reality of race as a social construct, which parallels the arguments mentioned above about settler colonialism within the sociology of race and ethnicity. This dynamic nature of race means that it can and does change, and is influenced by various social factors and processes at both the micro and macro level which is why a critical approach to the interpretation and contextualization of results from quantitative research is imperative to furthering understanding without sacrificing reflexivity (Zuberi & Bonilla-Silva, 2008). Ultimately, since the structures of racialization and settler-colonialism inform each other, an analysis of racial decision-making would be incomplete without considering how Indigenous racial formation came to be in the US.

## Theoretical Framework

In 2018 the Journal of Race Ethnicity and Education published a special issue titled "QuantCrit: Rectifying Quantitative Methods Through Critical Race Theory", where various scholars critically engaged with quantitative research and methods, using CRT to determine whether quantitative research is capable of being used to further a critical race agenda in education research and beyond (Garcia et al., 2018). This approach is not new, however, and has already been established in Du Bosian sociology, most notably in Du Bois’s The Philadelphia Negro and The Souls of Black Folk. The Philadelphia Negro has been described as a great example of an alternate approach to quantitative research that can be liberating and work to counter oppressive racial narratives while presenting statistical analysis alongside theoretical arguments (Conwell & Loughran, 2023). Du Bois’s approach came from necessity due to the popularity of theories like Social Darwinism which placed Black people at the bottom of the hierarchy, a reality that required the questioning of assumptions and working against the grain of how quantitative research was and often continues to be done. In The Souls of Black Folk Du Bois uses his lived experiences to theorize and explain what it was to be Black in American society. Emphasizing historical context and providing an explanation for the differences in perception between Black Americans and everyone else, most notably through double consciousness, the veil and the barriers of the color line. <!-- COMMENT: This is way too much. if QuantCrit undergrids your approach, then just show in it how you conduct the research. You do not need an entire paragraph exploring the connections between QuantCrit and Du Boisian sociology, because none of this has any direct bearing on the domain you are currently researching. -->  In this article, this approach will be replicated as best as possible, and part of that requires delving into settler colonial theory and current research related to Indigenous identity in the U.S. to create a historical and theoretical context that this paper will rely on to relate and explain its findings <!-- COMMENT: You don't need to make these arguments about why you are going to make other arguments! Just make the argument.-->.

In his article “Settler Colonialism and the Elimination of the Native,” Patrick Wolfe describes what he coined the ‘logic of elimination’ (2006). This logic explains the operation of colonizers arriving and settling on lands already inhabited by another group, specifically that settler colonialism requires the implementation of violence to achieve the goals of a settler society and replace Indigenous structures <!-- COMMENT: is "structures" a commonly used word here? I am not sure what you mean by it and you use structure in a different way below. I think its simpler to say the goal of settler society is indigenous removal. --> with those of the settler society. The replacement of Indigenous structures ultimately includes the removal and dispossession of Indigenous people themselves, which was justified through the racialization of Indigenous people which continues today, speaking to Wolfe’s assertion that settler colonialism is a “structure rather than an event” (Gonzales & Kertész, 2023; Wolfe, 2006, p. 390). Since the structures of racialization and settler-colonialism inform each other, an analysis of racial decision-making would be incomplete without considering how Indigenous racial formation came to be in the US. <!-- COMMENT: This last sentence is a duplicate of the last sentence of the previous section. --> <!-- COMMENT: Did you read Wolfe's Land, Labor, Difference article? That is the key Wolfe article to understand the relationship between settler colonialism and patterns of racial identification for those of indigenous ancestry. -->

For Indigenous Americans, membership within a tribe is often a key part of this process to distinguish between those who are Indigenous and those who are not. This concept is demonstrated in research done to identify what barriers to tribal enrollment exist for individuals who have been separated from their family and community via adoption and the foster system (Landers et al., 2021). Tribal enrollment was identified as key for individuals to establish and prove their own identity and seek collective identity within the tribe. Essentially, tribal enrollment exists as a “process of mutual verification between the individual and the tribe” (Landers et al., 2021, p. 1374). Additionally, the forceful removal of Indigenous children from their families between the 1800s and 1970s by the United States federal government–-via relocation, boarding schools, child welfare removal, adoption era practices, and so on, has created a group of Indigenous Americans, and their descendants, who have been removed from their family and culture thus further causing turmoil in how they conceptualize their identity <!-- CITATIONS -->. The Federal Acknowledgement Project of 1978, furthered this process since it created an administrative process that required any group <!-- COMMENT: Any indigenous group, correct? I think we are talking about the laws that effectively create blood quantum rules, but I am not clear --> wishing to be acknowledged by the federal government as a legitimate group to meet certain criteria established by the federal government, further establishing Indigenous American identity as an administrative process rather than of racial, ethnic, or cultural identity (Gonzales et al., 2007). The process explained above is one of the many ways the ‘elimination of the native’ has been accomplished, although elimination does not mean only the physical elimination of Indigenous people. For example, in their article “Settler Colonialism as eco-social structure and the production of colonial ecological violence,” Bacon emphasized that settler-colonial elimination projects are not just limited to traditional perceptions of genocide but include assimilation (2019) <!-- COMMENT: Nice to get in a UO grad, but Jules was not the first to make this claim by far! Patrick Wolfe's work certainly emphasizes it and likely others before him --> . The boarding schools Indigenous children were forced to attend not only utilized physical harm and violence but perpetuated colonial ecological violence by disrupting the passing on of traditional ecological knowledge from elder to child in favor of knowledge deemed superior by settlers (2019).

Blood quantum exists as a common requirement for tribal membership. The blood quantum minimum functions as a benchmark of authenticity for who is recognized as Indigenous by the tribal government and continues to endure as a key feature of Indigenous identity, and ultimately is an important part of determining who can access the few resources offered by the U.S. government such as healthcare and social services through the Indian Health Service (IHS) and the Bureau of Indian Affairs (BIA) (McKay, 2021; Rodriguez-Lonebear, 2021). This system of blood quantum was established as criteria for federal recognition in response to laws like Virginia’s Racial Integrity Act of 1924 and the Indian Reorganization Act (IRA) of 1934 which helped establish racial divisions and criteria on how to quantify Indigenous identity, respectively (Gonzales et al., 2007). The various laws related to blood quantum establish that determining Indigenous identity functions as an administrative process rather than a cultural process of a distinct racial and ethnic group, allowing a governing body to determine who can identify as Indigenous, the goal is to create an ever-shrinking population (Lambert, 2019). This concept demonstrated what McKay calls the American Indian Legal Identity (AILI). This idea is a proxy for an Indigenous identity borne of legal and federal policy, devoid of Indigeneity's cultural and community-based components (2013). <!-- COMMENT: Not sure I agree with your assessment of the role of these laws. I am not an expert, but I believe you need to distinguish historical usage from contemporary. Tribes now have ultimate control over tribal membership decisions and many tribes have established extremely low barriers to entry which has the opposite effect of shrinking populations. -->

Keeping the history of racial formation for Indigenous Americans in mind is important to be aware of when going into the results of this analysis, especially when considering how settler colonial theory can -both inform and support potential findings. Furthermore, the interpretation of results through the lens of CRT, QuantCrit, or settler-colonialism may not be able to account for every finding. However, engaging with the social reality multiracial Indigenous Americans navigate only serves to expand and explain current understandings regarding Indigenous identity, particularly multiracial Indigenous identity

<!-- COMMENT: This front end framing needs a good deal of restructuring. If I was a a reader not familiar with your work, I would be very confused at this point about what you were actually going to do and how it relates to any of the stuff you have been talking about above. You are going very into depth on specific readings or issues without linking them back to how they relate to your research question. You can cut a lot of the depth here in favor of greater breadth. Don't pick out one reading to be the focus of a paragraph, but rather write a paragraph that makes a point and bring multiple readings into it. I usually tell students to think of this front end framing as a funnel. You want to funnel the reader directly toward the importance of your research question. -->

## Methods and Data

### Sample Selection and Data

The data used for this project was compiled from the American Community Survey (ACS) data for 2010 - 2020. I seek to analyze the race reporting of individuals who indicate Indigenous ancestry in addition to another distinct ancestry group. On the ACS, respondents are asked about their ancestry via the question “What is this person's ancestry or ethnic origin,” and have two lines to list their responses which are manually typed and inputted as their responses. Respondents are also asked about their race and can mark one or more racial categories.

Before getting into the specifics of sample selection, there is a needed conversation about terminology. <!-- COMMENT: this is a throat-clearing sentence. You can delete it --> For this study, the term Indigenous is used in place of the American Indian and Alaskan Native (AIAN) category for simplicity's sake as it functions as a collective term since this project does not focus on a specific Nation, reservation, group, and so on. It should be noted that this is not done to imply Indigenous people are a monolith rather is a necessity when conducting a project that seeks to find patterns of racial identification that span across individual groups as evidence of the unique role of settler colonialism in the racialization of Indigenous Peoples in the U.S.

Part of the data organization of this project required the identification of a multiracial population, specifically an Indigenous multiracial population. First, any respondents that had an artificially allocated race response were removed, the goal was to make sure the race response was indicative of the respondents' true racial response. <!-- COMMENT: You are moving back and forth between describing the race and ancestry question in a way that will be hard for readers to undersand. --> Then the Goldstein and Morning method was adopted, which relies on looking at adults' self-reported ancestry <!-- CITE --> . This approach is beneficial since it does not rely on the race question to identify a multiracial Indigenous population, identifies a multiracial population that is aware of their ancestral descent and allows for a detailed response due to the free response nature of the question. Due to the free response style of the ancestry question, about 500 unique responses exist which require collapsing into specific racial ancestries <!-- COMMENT: its not completely "free" - the Census does remove or combine some responses -->. The U.S. Census follows the 1997 Office of Management and Budget (OMB) standards on race and ethnicity and lists the following groups: White, Black or African American, Asian, American Indian or Alaskan Native, and Native Hawaiian or Pacific Islander. It also includes the Hispanic or Latino ethnic category <!-- COMMENT: Not clear from the way you state that this that it is a separate question -->. For example, the Black ancestry group was created by combining all those whose ancestry fell under the groups of African-American, non-Latin Caribbean, and Sub-Saharan African. The Asian category included those whose ancestry fell under East, South East Asian, and South Asian. The White category included those with European, North African, and Middle Eastern ancestry. Respondents who listed their race as Latino or Pacific Islander were dropped from the potential multiracial Indigenous population <!-- COMMENT: You didn't drop Latino responses, its one of the variables! So I think you mean Latino ancestry responses? But there is nothing special about these two ancestries. You dropped anyone who didnt' fit a White/Indigenous, Black/Indigenous, or Asian/Indigenous ancestry - that is what you need to explain and justify. -->. The reason for dropping Latino is that it functions as an ethnic category on the Census/ACS and respondents of Latino or Hispanic origin can be any race, so Latino does not indicate a specific racial ancestry that is easily interpretable as distinct from Indigenous. The issue with the Pacific Islander category is that since they are Indigenous to their respective nations this could create the issue of a double Indigenous ancestry group depending on how respondents perceive their ancestry and to avoid this potential discrepancy they were also excluded from the sample.

After the final case selection was concluded the sample size was 490,172 respondents. Once this population was identified their ancestry was compared to their racial response. When reviewing respondents' racial ancestry and their race response, four distinct identification patterns were identified) Indigenous alone, which contained respondents that listed their race solely as Indigenous regardless of their multiracial ancestry 2) Consistent, which included respondents that acknowledged their racial ancestry in full by selecting a racial response that reflected a multiracial identity 3) Other Race Alone, where respondents' race response did not reflect their Indigenous ancestry and selected their identity solely as the other race 4) Inconsistent, where respondents race response did not reflect their racial ancestry in any capacity. The inconsistent group was dropped from further analysis, bringing the sample to 478,680 respondents. <!-- COMMENT: Needs to be discussed and justified, including the relatively large number of inconsistent Asian/Indigenous respondents -->.

### Analysis 

<!-- COMMENT: Analysis is an unusual title, maybe "Analytical Approach" -->

I use several multinomial logistic regression models to estimate the relationship between racial ancestry, race response, and the predictor variables. The dependent variable has three categories of race response: Indigenous Alone, Consistent (multiracial), and Other Race Alone. The key independent variable is racial ancestry. The predictor variables are metropolitan status (unknown status, non-metropolitan area, metropolitan area), homeland area (binary), educational attainment (less than HS diploma, HS diploma, some college, 4-year degree or more), gender (binary), Hispanic origin (binary), and age (continuous, 18+).

```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -2))+
  theme(axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0))
```

<!-- COMMENT: If you show a figure in the paper it should be described in the text. This is a good figure which shows some big differences in the outcome across your three ancestry cases, but you just leave it hanging. -->

## Results

### *Multinomial Logistic Regression Models*

My research question focuses on understanding the relationship between race response and racial ancestry <!-- COMMENT: Is that your research question? I thought it was more focused on the issue of rurality, homeland proximity, and education which is missing from the front end of the paper -->. I use the focal predictor variables to answer this question by identifying the relationship between the dependent and independent variables and the predictor variables. <!-- COMMENT: That doesn't answer your research question because racial ancestry is not a variable in your models but rather the backdrop for how you selected cases. Its not your research question! --> Each table presents coefficients, standard errors, and significance levels for the multinomial logistic regression models predicting race response with the predictor variables. The reference group separates these tables: White and Indigenous, Black and Indigenous, Asian and Indigenous. The model results compare the predicted race response of Indigenous Alone and Consistent to Other Race Alone with a separate table for each ancestry group. The base models of each table look at racial responses in relation to metropolitan status, the subsequent model adds Homeland status and the final model educational attainment, gender, Hispanic origin, and age. This setup creates three sets of tables with three models each for a total of 9 models.

### *Metropolitan Status*

Metropolitan status refers to whether respondents live in a metropolitan area or not. Living in a metropolitan area demonstrated a unique relationship between each group. For example, it was found that for respondents with White and Indigenous ancestry, the log odds of identifying as Indigenous Alone versus White while living in a metropolitan area were (-0.197) <!-- COMMENT: So, I prefer to keep log-odds in the tables but in the text I think it makes sense to convert these. So I would say the odds of identifying as Indigenous Alone versus White were 18% lower for those in metropolitan areas than non-metropolitan areas. -->. However, once Homeland status was accounted for it was found that the log odds of identifying as Indigenous Alone versus White in a metropolitan area is (0.187), indicating that living in a metropolitan area that is also a Homeland area is associated with an increased likelihood of an Indigenous identity. The log odds of respondents identifying as multiracial versus White when living in a metropolitan area were (0.158). Once Homeland status is accounted for, living in a metropolitan area increases log odds of identifying as multiracial versus White (0.348).

For respondents with Black and Indigenous ancestry, living in a metropolitan area has a negative association with identifying as Indigenous versus Black (0.708), however, there is a positive albeit non-significant relationship between metropolitan status and multiracial identity. Once Homeland status is accounted for this relationship flipped. Living in a metropolitan area has a non-significant but negatively associated relationship with identifying as Indigenous versus Black (-0.206). A significant but positive relationship exists between living in a metropolitan area and identifying as multiracial versus Black (0.265).

For respondents with Asian and Indigenous ancestry, living in a metropolitan area was negatively associated with identifying as Indigenous versus Asian (-1.769) as well as identifying as multiracial versus Asian (-0.550). Once Homeland status was accounted for there continued to be a negative association between identifying as Indigenous versus Asian (-0.787) and multiracial versus Asian however the relationship was not significant (-0.103).

Accounting for education, gender, Hispanic origin, and age led to minimal changes across all three ancestry groups.

### *Homeland Status*

Living in a location with Homeland status means that a respondent’s household is in an area that includes an area that has been designated as an American Indian, Alaska Native, or Native Hawaiian (AIANNH) area (Bureau, n.d.). For respondents with White and Indigenous ancestry, living on a Homeland was positively associated with identifying as Indigenous (1.452) and multiracial (0.799) versus identifying as White. For respondents with Black and Indigenous ancestry living on a Homeland is positively associated with an Indigenous identity versus Black (1.177) and the same holds for a multiracial identity versus Black although not to the same degree (0.509). The Homeland status effect was strongest for respondents with Asian and Indigenous ancestry, as living in a Homeland was associated with an increased likelihood to identify as Indigenous versus Asian(1.900), the same holds for identifying as multiracial versus Asian (0.843).

Accounting for education, gender, Hispanic origin, and age led to minimal changes across all three ancestry groups

### *Educational Attainment*

Educational attainment consistently has a significant relationship with race response across ancestry groups, however, three unique patterns become apparent once a college degree is obtained. For those with White and Indigenous ancestry obtaining a college degree pushes respondents away from Whiteness as it is associated with an increase in the log odds of identifying as Indigenous alone versus White at (0.372) and Multiracial versus White at (0.328) log odds. For those with Black and Indigenous ancestry obtaining a college degree pushes respondents away from Indigenous Alone and towards a Multiracial identity at (-0.355) and (0.144) log odds respectively. For those with Asian and Indigenous ancestry obtaining a college degree signaled a move away from identifying as Indigenous Alone and towards identifying as Asian alone with the log odds of identifying as Indigenous Alone being (-1.099).

### *Gender*

Gender demonstrated a unique relationship amongst each of the ancestry groups. For example, for respondents of White and Indigenous ancestry it was found that men were more likely to identify as Indigenous or multiracial versus White at similar rates (0.259 v. 0.255). When a respondent has Black and Indigenous ancestry, it was found that gender did not have a significant relationship between identifying as Indigenous versus Black but it was found that men are more likley to identify as multiracial versus identifying as Black than women (0.248). The opposite effect was seen with those that have Asian and Indigenous ancestry, as men were found to be less likely to identify as multiracail versus as Asian than women (-0.300).

### *Hispanic Origin*

In the ACS, Hispanic Origin refers to individuals of Hispanic/Spanish/Latino origin, where origin is defined by the Census Bureau as ancestry, lineage, heritage, nationality group, or country of birth (Bureau, n.d.). Additionally, those of Hispanic Origin can be any race or any combination of races. In the results of this study, Hispanic origin is associated with a pattern of increased attachment towards an Indigenous identity and multiraciality for all ancestry groups. Respondents who listed White and Indigenous ancestry and are of Hispanic Origin have increased log odds of identifying as Indigenous Alone versus White ( 2.161), and the log odds of identifying as Multiracial versus White are also positive but diminished compared to Indigenous alone (1.503). Black and Indigenous respondents of Hispanic Origin have increased log odds of identifying as Indigenous Alone versus Black (1.691) and similar log odds of identifying as Multriacal versus Black alone (1.405). Respondents of Asian and Indigenous ancestry who are also of Hispanic origin have positive log odds of identifying as Indigenous versus Asian (1.092) as well as multiracial versus Asian (0.804). Ultimately Hispanic origin increases the likelihood of identifying as Indigenous or multiracial, yet the effect is strongest to weakest for those with White and Indigenous ancestry, those with Black and Indigenous ancestry, and Asian and Indigenous ancestry, respectively.

### *Age*

Age as a predictor variable demonstrated minimal effects on racial response, and significant results were only observed for two ancestry groups. Amongst respondents with White and Indigenous ancestry it was observed that for every year in age there was a minimal increase in the likelihood for respondents to identify as Indigenous versus White (0.002). The effect was similarly minimal for multiracial versus White (0.001).

<!-- COMMENT: I see why you were having trouble with the results section, but its not the results section that is the problem. The problem is that you didn't adequately frame the research questions of the paper in the front end so that by the time we get here you just have ... stuff ... to report. You haven't established why we should care about any of this or what expectations we might have about it. I thought the original focus of the paper was on the spatial distribution issue, particularly related to proximity to homelands. That is now completely missing. -->

## Discussion

The previously discussed results demonstrated patterns that both reaffirmed previous findings related to Indigenous identity but also revealed unique deviations. After further reviewing metropolitan status three distinct patterns were identified concerning racial identity responses amongst the three mixed racial ancestry groups.

First, for respondents with both White and Indigenous ancestry, there was a noticeable shift away from a White identity towards embracing an Indigenous or a multiracial identity.Second, for those with Black and Indigenous ancestry, living in a metropolitan area was associated with a different pattern. These respondents tended to move away from an Indigenous identity and towards identifying as Black, while multirciality remained an option at a similar rate of association. Finally, for respondents with Asian and Indigenous ancestry, metropolitan status correlated with a tendency to move away from identifying with their Indigenous ancestry and towards identifying exclusively as Asian rather than multiracial or Indigenous alone.

<!-- COMMENT: All of the stuff here at the end is what should be up in the front -->

-   living in urban spaces is associated with the creation of a unique identity that does not exist without tension (Malika et al., 2024).

-   Discuss urban migration and relation to settler colonialism (Rosenthal, 2012)

-   Tensions of multiracial identity and assertion of agency through racial decision making (selection one option or another or choosing both) (Cardwell et al., 2020)

Homeland Status

-   Homeland status increased likelihood to identify as Indigenous for all groups, in order of effect WI, AI, and BI, also movement towards multiracial

-   Liebler Liebler Liebler

    -   Homelands and Indigenous Identity

    -   History, place, and racial self-representation in 21st century America

    -   Challenge with the discussion of tribal sovereignty in enrollment

-   Native DNA, relate to sovereignty, relations, connection, enrollment rather than focusing on determining authenticity due to blood quantum standards ie churning race liebler article

    -   The blood line: racialized boundary making (Rodriguez-Lonebear, 2021)

    -   Native American DNA (TallBear, 2013)

    -   Maybe this can connect to sovereignty: Tribal Sovereignty and the Limits of Race for American Indians (Small-Rodriguez & Rocha Beardall, 2023)

Hispanic Origin

-   Latinization of indigenous identity (Campbell-Montalvo, 2023)

-   The distinction between indigenous and Latino/Hispanic identity (Ybarra, 2023)

    -   Why do some keep it one but not the other?

-   How migration studies can reproduce the logic of elimination, work against the idea that when Indigenous people travel they are no longer indigenous as explanation for hispanic origin and indigeneity relationship (Blackwell et al., 2017)

Educational Attainment

-   College education pushed WI away from a White identity, for BI pushed towards multiracial identity and Black identity, for AI college education pushed towards Asian identity

    -   Potential identity tension?

-   Escbach found that people with higher level of education found Indigenous identity more appealing, especially in metropolitan areas (Eschbach et al., 1998)

-   “With the exception of Black Native Americans, all of the Native American male groups have lower levels of schooling”; “With the possible exception of Black Native American women, all groups of Native American men and women have some sort of disadvantage in terms of schooling or earnings” (Huyser et al., 2010)

    -   Seems consistent with findings for BI respondents across gender

-   AI is where I am most interested and wondering how much identity conflict/tension exists

    -   Are respondents choosing race that matches with stereotypes? I feel like that is too simple of an answer, perhaps perception of perpetual foreigner leads to more attachment to Asian identity?

    -   Racial and ethnic stratification in educational achievement and attainment (Kao & Thompson, 2003)

    -   “The second contemporary pattern of multiracial identification we discern might be called simply the ‘‘new’’ or even ‘‘immigrant’’ regime whereby a multiple-race label is selected. This convention seems to dominate among all the groups of partial Asian ancestry” (Gullickson & Morning, 2011), this continues to be true as can be seen in Figure 1 (i haven’t named them yet lol)

Gender

-   I need to look more into research between multiraciality and gender, indigenous identity and gender, etc

-   Not sure how to approach yet

Age

-   Age did not have major effect, however thinking in cohorts should be interesting

-   WI group saw increase to both multiracial and Indigenous identity with each year in age

-   BI only sees negative association with multiraciaity versus identifying just as Black

    -   One drop rule having more salience for older cohort/groups?

## References
