---
title: "Navigating Ancestry and Racial Classification Amongst Multiracial Indigenous American"
shorttitle: "Navigating Ancestry and Racial Classification"
abstract: Current research on race and multiracial individuals recognizes that Indigenous racial identity is fluid and often contested. Using a settler-colonial theoretical framework and the recommendations of QuantCrit literature, this research project expands understanding of multi-racial Indigenous identity and demonstrates how ancestry influences racial “decision-making” for Indigenous Americans. By leveraging data from the American Community Survey (ACS) from 2010 to 2020, the relationship between an individuals reported ancestry and their self-identified racial classification is investigated by estimating multinomial logistic regression models. The results indicate the relationship between various predictorss, and a persons likelihood to identify as Indigenous alone, multi-racial, or to distance themself from their Indigenous ancestry and identity in favor of their other racial identity. When these findings are evaluated within the context of settler-colonialism, many previously confounding findings can be linked to the social reality that respondents navigate as Indigenous Americans.
keywords: [Indigenous identity, multiracial studies, ancestry, racial decision-making, racial classification, settler-colonialism, QuantCrit, American Community Survey]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/apa-6th-edition.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData"))
```

## Introduction

Who is Indigenous and who is not? Where do multiracial Indigenous Americans fit in that equation? One group of scholars has asked similar questions in an attempt to better understand Indigenous identity by determining which factors are associated with strong Indigenous identity, how this identity is contested and understood at an interpersonal and community level, and how the Indigenous population has grown over time (Liebler, 2010; Lucero, 2014; McKay, 2021). Indigenous identity becomes even more complicated when considering multiracial Indigenous Americans, which has led to additional research in this area, although limited. Articles on Indigenous identity generally cover the racial decision-making process for multiracial Indigenous people in the United States, since many Indigenous people have mixed racial ancestries. This social reality requires Indigenous American respondents to make conscious decisions when listing their race. In the case of multiracial Indigenous people their known racial ancestry either conflicts with or affirms their answer, and they must reconcile with that decision. Understanding this decision-making process is crucial for gaining insights into contemporary shifts in racial boundary formation and the individual considerations that shape racial identification for Indigenous Americans.

Dwanna L. McKay, Kirsten Vinyeta, and Kari M. Norgaard published an article titled “Theorizing Race and Settler Colonialism within U.S. Sociology” which calls for a critical approach to how race is studied within Sociology, particularly how the focus on assimilation fails to confront its violent and often deadly nature for Indigenous Americans under settler colonialism (2020). Their article established that as researchers and academics, we are complicit in the erasure of Indigenous people, and the continued reification of harmful narratives can only be stopped by engaging with settler colonialism theory when conducting our research.

These considerations led to the theoretical engagement explored later in this paper to prepare for a critical and nuanced approach to quantitative research that has often been critiqued for its inability to speak to individual experiences. For those who study race, there have been criticisms of using race as a static category that does not reflect the social reality of race as a social construct, which parallels the arguments mentioned above about settler colonialism within the sociology of race and ethnicity. This dynamic nature of race means that it can and does change, and is influenced by various social factors and processes at both the micro and macro level which is why a critical approach to the interpretation and contextualization of results from quantitative research is imperative to furthering understanding without sacrificing reflexivity (Zuberi & Bonilla-Silva, 2008). Ultimately, since the structures of racialization and settler-colonialism inform each other, an analysis of racial decision-making would be incomplete without considering how Indigenous racial formation came to be in the US.

## Theoretical Framework

In 2018 the Journal of Race Ethnicity and Education published a special issue titled "QuantCrit: Rectifying Quantitative Methods Through Critical Race Theory", where various scholars critically engaged with quantitative research and methods, using CRT to determine whether quantitative research is capable of being used to further a critical race agenda in education research and beyond (Garcia et al., 2018). This approach is not new, however, and has already been established in Du Bosian sociology, most notably in Du Bois’s The Philadelphia Negro. The Philadelphia Negro has been described as a great example of an alternate approach to quantitative research that can be liberating and work to counter oppressive racial narratives while presenting statistical analysis alongside theoretical arguments (Conwell & Loughran, 2023). Du Bois’s approach came from necessity due to the popularity of theories like Social Darwinism which placed Black people at the bottom of the hierarchy, a reality that required the questioning of assumptions and working against the grain of how quantitative research was and often continues to be done. In this article, this approach will be replicated as best as possible, and part of that requires delving into settler colonial theory and current research related to Indigenous identity in the U.S. to create a historical and theoretical context that this paper will rely on.

In his article “Settler Colonialism and the Elimination of the Native,” Patrick Wolfe describes what he coined the ‘logic of elimination’ (2006). This logic explains the operation of colonizers arriving and settling on lands already inhabited by another group, specifically that settler colonialism requires the implementation of violence to achieve the goals of a settler society and replace Indigenous structures with those of the settler society. The replacement of Indigenous structures ultimately includes the removal and dispossession of Indigenous people themself, which was justified through the racialization of Indigenous people (Gonzales & Kertész, 2023). Since the structures of racialization and settler-colonialism inform each other, an analysis of racial decision-making would be incomplete without considering how Indigenous racial formation came to be in the US.

For Indigenous Americans, membership within a tribe is often a key part of this process to distinguish between those who are Indigenous and those who are not. This concept is demonstrated in research done to identify what barriers to tribal enrollment exist for individuals who have been separated from their family and community via adoption and the foster system (Landers et al., 2021). Tribal enrollment was identified as key for individuals to establish and prove their own identity and seek collective identity within the tribe. Essentially, tribal enrollment exists as a “process of mutual verification between the individual and the tribe” (Landers et al., 2021, p. 1374).  Additionally, the forceful removal of Indigenous children from their families between the 1800s and 1970s by the United States federal government–-via relocation, boarding schools, child welfare removal, adoption era practices, and so on, has created a group of Indigenous Americans, and their descendants, who have been removed from their family and culture thus further causing turmoil in how they conceptualize their identity. The Federal Acknowledgement Project of 1978, furthered this process since it created an administrative process that required any group wishing to be acknowledged by the federal government as a legitimate group to meet certain criteria established by the federal government, further establishing Indigenous American identity as an administrative process rather than of racial, ethnic, or cultural identity (Gonzales et al., 2007). The process explained above is one of the many ways the ‘elimination of the native’ has been accomplished, although elimination does not mean only the physical elimination of Indigenous people. For example, in their article “Settler Colonialism as eco-social structure and the production of colonial ecological violence,” Bacon emphasized that settler-colonial elimination projects are not just limited to traditional perceptions of genocide but include assimilation (2019). The boarding schools Indigenous children were forced to attend not only utilized physical harm and violence but perpetuated colonial ecological violence by disrupting the passing on of traditional ecological knowledge from elder to child in favor of knowledge deemed superior by settlers (2019).

Blood quantum exists as a common requirement for tribal membership. The blood quantum minimum functions as a benchmark of authenticity for who is recognized as Indigenous by the tribal government and continues to endure as a key feature of Indigenous identity, and ultimately is an important part of determining who can access the few resources offered by the U.S. government (McKay, 2021; Rodriguez-Lonebear, 2021). This system of blood quantum was established as criteria for federal recognition in response to laws like Virginia’s Racial Integrity Act of 1924 and the Indian Reorganization Act (IRA) of 1934 which helped establish racial divisions and criteria on how to quantify Indigenous identity, respectively (Gonzales et al., 2007). The various laws related to blood quantum establish that determining Indigenous identity functions as an administrative process rather than a cultural process of a distinct racial and ethnic group, allowing a governing body to determine who can identify as Indigenous, the goal is to create an ever-shrinking population (Lambert, 2019). This concept demonstrated what McKay calls the American Indian Legal Identity (AILI). This idea is a proxy for an Indigenous identity borne of legal and federal policy, devoid of Indigeneity's cultural and community-based components (2013). 

Keeping the history of racial formation for Indigenous Americans in mind is important to be aware of when going into the results of this analysis, especially when considering how settler colonial theory can -both inform and support potential findings. Furthermore, the interpretation of results through the lens of CRT, QuantCrit, or settler-colonialism may not be able to account for every finding. However, engaging with the social reality multiracial Indigenous Americans navigate only serves to expand and explain current understandings regarding Indigenous identity, particularly multiracial Indigenous identity

## Methods and Data

### *Sample Selection and Data*

The data used for this project was compiled from the American Community Survey (ACS) data for 2010 - 2020.  I seek to analyze the race reporting of individuals who indicate Indigenous ancestry in addition to another distinct ancestry group. On the ACS, respondents are asked about their ancestry via the question “What is this person's ancestry or ethnic origin,” and have two lines to list their responses which are manually typed and inputted as their responses. Respondents are also asked about their race and can mark one or more racial categories

Before getting into the specifics of sample selection, there is a needed conversation about terminology. For this study, the term Indigenous is used in place of the American Indian and Alaskan Native (AIAN) category for simplicity's sake as it functions as a collective term since this project does not focus on a specific Nation, reservation, group, and so on. It should be noted that this is not done to imply Indigenous people are a monolith rather is a necessity when conducting a project that seeks to find patterns of racial identification that span across individual groups as evidence of the unique role of settler colonialism in the racialization of Indigenous Peoples in the U.S.

Part of the data organization of this project required the identification of a multiracial population, specifically an Indigenous multiracial population. First, any respondents that had an artificially allocated race response were removed, the goal was to make sure the race response was indicative of the respondents' true racial response. Then the Goldstein and Morning method was adopted, which relies on looking at adults' self-reported ancestry. This approach is beneficial since it does not rely on the race question to identify a multiracial Indigenous population, identifies a multiracial population that is aware of their ancestral descent and allows for a detailed response due to the free response nature of the question. Due to the free response style of the ancestry question, about 500 unique responses exist which require collapsing into specific racial ancestries. The U.S. Census follows the 1997 Office of Management and Budget (OMB) standards on race and ethnicity and lists the following groups: White, Black or African American, Asian, American Indian or Alaskan Native, and Native Hawaiian or Pacific Islander. It also includes the Hispanic or Latino ethnic category. For example, the Black ancestry group was created by combining all those whose ancestry fell under the groups of African-American, non-Latin Caribbean, and Sub-Saharan African. The Asian category included those whose ancestry fell under East, South East Asian, and South Asian. The White category included those with European, North African, and Middle Eastern ancestry. Respondents who listed their race as Latino or Pacific Islander were dropped from the potential multiracial Indigenous population. The reason for dropping Latino is that it functions as an ethnic category on the Census/ACS and respondents of Latino or Hispanic origin can be any race, so Latino does not indicate a specific racial ancestry that is easily interpretable as distinct from Indigenous. The issue with the Pacific Islander category is that since they are Indigenous to their respective nations this could create the issue of a double Indigenous ancestry group depending on how respondents perceive their ancestry and to avoid this potential discrepancy they were also excluded from the sample.

After the final case selection was concluded the sample size was 490,172 respondents. Once this population was identified their ancestry was compared to their racial response. When reviewing respondents' racial ancestry and their race response, four distinct identification patterns were identified) Indigenous alone, which contained respondents that listed their race solely as Indigenous regardless of their multiracial ancestry 2) Consistent, which included respondents that acknowledged their racial ancestry in full by selecting a racial response that reflected a multiracial identity 3) Other Race Alone, where respondents' race response did not reflect their Indigenous ancestry and selected their identity solely as the other race 4) Inconsistent, where respondents race response did not reflect their racial ancestry in any capacity. The inconsistent group was dropped from further analysis, bringing the sample to 478,680 respondents.

### *Analysis*

I use several multinomial logistic regression models to estimate the relationship between racial ancestry, race response, and the predictor variables. The dependent variable has three categories of race response: Indigenous Alone, Consistent (multiracial), and Other Race Alone. The key independent variable is racial ancestry. The predictor variables are metropolitan status (unknown status, non-metropolitan area, metropolitan area), homeland area (binary), educational attainment (less than HS diploma, HS diploma, some college, 4-year degree or more), gender (binary), Hispanic origin (binary), and age (continuous, 18+).

```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -2))+
  theme(axis.text.x = element_text(angle = -30, vjust = 1, hjust = 0))
```

```{r}
```

```{r}
#| label: tbl-white-indigenous
#| tbl-cap: White/Indigenous Models
#| results: asis
model1 <- multinom(race_response~MetroC,
                  data=subset(DATA, race_ancestry=="Indigenous/White"))
model2 <- update(model1, .~.+HomelandP)
model3 <- update(model2, .~.+EducA+Gender+Hispanic+Age)
modelsummary(list(model1,model2,model3),
             digits=3, shape = term ~ model+response, stars = TRUE)
```

## Results

### *Multinomial Logistic Regression Models*

My research question focuses on understanding the relationship between race response and racial ancestry. I use the focal predictor variables to answer this question by identifying the relationship between the dependent and independent variables and the predictor variables. Each table presents coefficients, standard errors, and significance levels for the multinomial logistic regression models predicting race response with the predictor variables. The reference group separates these tables: White and Indigenous, Black and Indigenous, Asian and Indigenous. The model results compare the predicted race response of Indigenous Alone and Consistent to Other Race Alone with a separate table for each ancestry group. The base models of each table look at racial responses in relation to metropolitan status, the subsequent model adds Homeland status and the final model educational attainment, gender, Hispanic origin, and age. This setup creates three sets of tables with three models each for a total of 9 models total.

### *Metropolitan Status*

Metropolitan status refers to whether respondents live in a metropolitan area or not. Living in a metropolitan area demonstrated a unique relationship between each group. For example, it was found that for respondents with White and Indigenous ancestry, the log odds of identifying as Indigenous Alone versus White while living in a metropolitan area were (-0.197). However, once Homeland status was accounted for it was found that the log odds of identifying as Indigenous Alone versus White in a metropolitan area is (0.187), indicating that living in a metropolitan area that is also a Homeland area is associated with an increased likelihood of an Indigenous identity. The log odds of respondents identifying as multiracial versus White when living in a metropolitan area were (0.158). Once Homeland status is accounted for, living in a metropolitan area increases log odds of identifying as multiracial versus White (0.348).

For respondents with Black and Indigenous ancestry, living in a metropolitan area has a negative association with identifying as Indigenous versus Black (0.708), however, there is a positive albeit non-significant relationship between metropolitan status and multiracial identity. Once Homeland status is accounted for this relationship flipped. Living in a metropolitan area has a non-significant but negatively associated relationship with identifying as Indigenous versus Black (-0.206). A significant but positive relationship exists between living in a metropolitan area and identifying as multiracial versus Black (0.265).

For respondents with Asian and Indigenous ancestry, living in a metropolitan area was negatively associated with identifying as Indigenous versus Asian (-1.769) as well as identifying as multiracial versus Asian (-0.550). Once Homeland status was accounted for there continued to be a negative association between identifying as Indigenous versus Asian (-0.787) and multiracial versus Asian however the relationship was not significant (-0.103).

Accounting for education, gender, Hispanic origin, and age led to minimal changes across all three ancestry groups.

### *Homeland Status*

Living in a location with Homeland status means that a respondent’s household is in an area that includes an area that has been designated as an American Indian, Alaska Native, or Native Hawaiian (AIANNH) area (Bureau, n.d.). For respondents with White and Indigenous ancestry, living on a Homeland was positively associated with identifying as Indigenous (1.452) and multiracial (0.799) versus identifying as White. For respondents with Black and Indigenous ancestry living on a Homeland is positively associated with an Indigenous identity versus Black (1.177) and the same holds for a multiracial identity versus Black although not to the same degree (0.509).

Accounting for education, gender, Hispanic origin, and age led to minimal changes across all three ancestry groups

### *Educational Attainment*

Educational attainment consistently has a significant relationship with race response across ancestry groups, however, three unique patterns become apparent once a college degree is obtained. For those with White and Indigenous ancestry obtaining a college degree pushes respondents away from Whiteness as it is associated with an increase in the log odds of identifying as Indigenous alone versus White at (0.372) and Multiracial versus White at (0.328) log odds. For those with Black and Indigenous ancestry obtaining a college degree pushes respondents away from Indigenous Alone and towards a Multiracial identity at (-0.355) and (0.144) log odds respectively. For those with Asian and Indigenous ancestry obtaining a college degree signaled a move away from identifying as Indigenous Alone and towards identifying as Asian alone with the log odds of identifying as Indigenous Alone being (-1.099).

### *Gender*

### *Hispanic Origin*

In the ACS, Hispanic Origin refers to individuals of Hispanic/Spanish/Latino origin, where origin is defined by the Census Bureau as ancestry, lineage, heritage, nationality group, or country of birth (Bureau, n.d.). Additionally, those of Hispanic Origin can be any race or any combination of races. In the results of this study, Hispanic origin is associated with a pattern of increased attachment towards an Indigenous identity and multiraciality for all ancestry groups. Respondents who listed White and Indigenous ancestry and are of Hispanic Origin have increased log odds of identifying as Indigenous Alone versus White ( 2.161), and the log odds of identifying as Multiracial versus White are also positive but diminished compared to Indigenous alone (1.503). Black and Indigenous respondents of Hispanic Origin have increased log odds of identifying as Indigenous Alone versus Black (1.691) and similar log odds of identifying as Multriacal versus Black alone (1.405). Respondents of Asian and Indigenous ancestry who are also of Hispanic origin have positive log odds of identifying as Indigenous versus Asian (1.092) as well as multiracial versus Asian (0.804). Ultimately Hispanic origin increases the likelihood of identifying as Indigenous or multiracial, yet the effect is strongest to weakest for those with White and Indigenous ancestry, those with Black and Indigenous ancestry, and Asian and Indigenous ancestry, respectively.

### *Age*

## Discussion

## Conclusion

## References
