---
title: "The Nexus Between Ancestry and Racial Classification Amongst Indigenous Americans"
shorttitle: "Paper"
abstract: Current research on race and multiracial individuals identifies that Indigenous racial identity is particularly fluid and often contested. Utilizing a settler-colonial theoretical framework, this research project aims to create a better understanding of Indigenous identity and answer how ancestry influences racial “decision-making” for Indigenous Americans. By leveraging census data from 2010-2020, the intricate relationship between individuals reported ancestry and their self-identified racial classification is investigated by employing various statistical analysis techniques to discern patterns, trends, and discrepancies in how these individuals perceive their ancestral heritage and racial identity.
keywords: [Indigenous, ancestery, race, identity]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/apa-6th-edition.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData"))
```

# Introduction

Indigenous Americans are a group that has historically been difficult to study due to the lasting and continuous effects of settler-colonialism examples include but are not limited to genocide, assimilation practices/policies, political disenfranchisement, and erasure @gullickson_choosing_2011 @liebler_joining_2016 @korgen_multiracial_2010 @snipp_american_1989 @bacon_settler_2019.

# Background

Settler-colonial scholars have recently urged for the inclusion of settler-colonialism when studying the historical and contemporary racial structure of the U.S @mckay_theorizing_2020

# Data and Methods

```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -2))
```

```{r}
#| label: tbl-multi-models
#| tbl-cap: Multinomial Models
#| results: asis
model1 <- multinom(race_response~HomelandP,
                  data=subset(DATA, race_ancestry=="Indigenous/White"))
model2 <- multinom(race_response~HomelandP+EducA,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
model3 <- multinom(race_response~HomelandP+EducA+Gender,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
model4 <- multinom(race_response~HomelandP+EducA+Gender+Hispanic,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
model5 <- multinom(race_response~HomelandP+EducA+Gender+Hispanic+MetroC,
                   data = subset(DATA, race_ancestry=="Indigenous/White"))
texreg(list(model1,model2,model3,model4,model5),
          digits=3)
```

# Results

In @fig-race-ancestry 72% of respondents who list Indigenous and Asian ancestry did not self-identify as Indigenous or Asian in regards to their racial identity. \^[Does this challenge previous research that suggests that those with part Asian ancestry tend to favor a multiracial identity @gullickson_choosing_2011. This can help lead to some future analysis and inform the inclusion of more literature on multiracial identity among mixed-race Asians]

# Conclusion
