---
title: "The Nexus Between Ancestry and Racial Classification Amongst Indigenous Americans"
shorttitle: "Paper"
abstract: Current research on race and multiracial individuals identifies that Indigenous racial identity is particularly fluid and often contested. Utilizing a settler-colonial theoretical framework, this research project aims to create a better understanding of Indigenous identity and answer how ancestry influences racial “decision-making” for Indigenous Americans. By leveraging census data from 2010-2020, the intricate relationship between individuals reported ancestry and their self-identified racial classification is investigated by employing multinomial logistic regression models to delineate the relationship between various predictors and the likelihood to identify as Indigenous alone, multi-racial or to distance and ultimately reject their Indigenous ancestry and identity in favor of their other racial identity.
keywords: [Indigenous, ancestery, race, identity]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/apa-6th-edition.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData"))
```

# Introduction

Indigenous Americans are a group that has historically been difficult to study due to the lasting and continuous effects of settler-colonialism examples include but are not limited to genocide, assimilation practices/policies, political disenfranchisement, and erasure @gullickson_choosing_2011 @liebler_joining_2016 @korgen_multiracial_2010 @snipp_american_1989 @bacon_settler_2019.

# Theoretical Framework

Settler-colonial scholars have recently urged for the inclusion of settler-colonialism when studying the historical and contemporary racial structure of the U.S @mckay_theorizing_2020

Quantcrit, neccesity for acknowledging that how we interpet numbers is influenced by our ontolgical and epistemological lens

# Methods and data
## Sample Selection and Data
The data used for this project was compiled from American Community Survey data from 2010-2020.I seek to analyze the race reporting of individuals who indicate Indigenous (define in some capacity? Footnote maybe) ancestry in addition to White, Black, or Asian ancestry.Individuals who identified as Latino or Pacific Islanders were not included in the sample due to either inconsistencies of.... or a desire to focus on Indigenous people of the continental United States.
Outcome variable: race response
Predictor variables: metro, homeland, education above high school, gender and hispanicity 
These ancestry responses were compared to what their race response and divided into the following groups: Indigenous alone, Multiracial, and Other Race alone where Indigenous alone consisted of those respondents who identified a mixed-ancestry of Indigenous plus one other racial ancestry yet listed their race response as Indigenous alone. Multiracial was those respondents who identified their mixed--Indigenous ancestry and listed a multiracial race response. Other race alone consisted of those respondents with a mixed-Indigenous racial ancestry but listed their racial response as the other race (ie. ancestry listed as Indigenous and Asian but list race as Asian only). 

## Methods
The analysis was conducted through creation of several multinomial logit models, which predict the likelihood of a respondent to identify as either Indigenous alone, Multiracial, or other race alone. Three sets of models were created, each set had a different reference category: "Idigenous/White", "Indigenous/Black", and "Indigenous/Asian"
```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -2))
```

```{r}
```

# Results


# Conclusion
