---
title: "Navigating Ancestry and Racial Classification Amongst Multiracial Indigenous American"
shorttitle: "Navigating Ancestry and Racial Classification"
abstract: Current research on race and multiracial individuals recognizes that Indigenous racial identity is fluid and often contested. Using a settler-colonial theoretical framework and the recommendations of QuantCrit literature, this research project expands understanding of multi-racial Indigenous identity and demonstrates how ancestry influences racial “decision-making” for Indigenous Americans. By leveraging data from the American Community Survey (ACS) from 2010 to 2020, the relationship between an individuals reported ancestry and their self-identified racial classification is investigated by estimating multinomial logistic regression models. The results indicate the relationship between various predictorss, and a persons likelihood to identify as Indigenous alone, multi-racial, or to distance themself from their Indigenous ancestry and identity in favor of their other racial identity. When these findings are evaluated within the context of settler-colonialism, many previously confounding findings can be linked to the social reality that respondents navigate as Indigenous Americans.
keywords: [Indigenous identity, multiracial studies, ancestry, racial decision-making, racial classification, settler-colonialism, QuantCrit, American Community Survey]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/apa-6th-edition.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data","data_constructed","analytical_data.RData"))
```

# Introduction

Who is Indigenous and who is not? Where do multiracial Indigenous Americans fit in this equation? One group of scholars has attempted to better understand Indigenous identity by asking similar questions by determining which factors are associated with strong Indigenous identity, how this identity is contested and understood at an interpersonal and community level, and how the Indigenous population has grown over time \[@liebler_homelands_2010 @lucero_its_2014 @mckay_real_2021. Indigenous identity becomes even more complicated when considering multiracial Indigenous Americans, which has led to additional research in this area, although limited. Articles on Indigenous identity generally cover the racial decision-making process for multiracial Indigenous people in the United States, since many Indigenous people have mixed racial ancestries. This social reality requires Indigenous American respondents to make conscious decisions when listing their race. In the case of multiracial Indigenous people their known racial ancestry either conflicts with or affirms their answer, and they must reconcile with that decision. Understanding this decision-making process is crucial for gaining insights into contemporary shifts in racial boundary formation and the individual considerations that shape racial identification for Indigenous Americans.

However, in 2020 Dwanna L. McKay, Kirsten Vinyeta, and Kari M. Norgaard published an article titled “Theorizing Race and Settler Colonialism within U.S. Sociology” which calls for a critical approach to how race is studied within Sociology, particularly how the focus on assimilation fails to confront its violent and often deadly nature of assimilation for Indigenous Americans under settler colonialism. Their article established that as researchers and academics, we are complicit in the erasure of Indigenous people, and the continued reification of harmful narratives can only be stopped by engaging with settler colonialism theory when conducting our research.

These considerations led to the theoretical engagement explored later in this paper to prepare for a critical and nuanced approach to quantitative research has often been critiqued for its inability to speak to individual experiences, and for those who study race, there have been critiques of using race as a static category that does not reflect the social reality of race as a social construct, which parallels the arguments mentioned above about settler colonialism within the sociology of race and ethnicity. This dynamic nature of race means that it can and does change, and is influenced by various social factors and processes at both the micro and macro level which is why a critical approach to the interpretation and contextualization of results from quantitative research is imperative to furthering understanding without sacrificing reflexivity @zuberi_white_2008.

# Theoretical Framework

In 2018 the Journal of Race Ethnicity and Education published a special issue titled "QuantCrit: Rectifying Quantitative Methods Through Critical Race Theory", where various scholars critically engaged with quantitative research and methods, using CRT to determine whether quantitative research is capable of being used to further a critical race agenda in education research and beyond @garcia_quantcrit_2018. This approach is not new, however, and has already been established in Du Bosian sociology, most notably in Du Bois’s The Philadelphia Negro. The Philadelphia Negro has been described as a great example of an alternate approach to quantitative research that can be liberating and work to counter oppressive racial narratives while presenting statistical analysis alongside theoretical arguments @conwell_quantitative_2023. Du Bois’s approach came from necessity due to the popularity of theories like Social Darwinism which placed Black people at the bottom of the hierarchy, a reality that required the questioning of assumptions and working against the grain of how quantitative research was and often continues to be done. In this article, this approach will be replicated as best as possible, and part of that requires delving into settler colonial theory and current research related to Indigenous identity in the U.S. to create a historical and theoretical context that this paper will rely on.

In his article “Settler Colonialism and the Elimination of the Native,” Patrick Wolfe describes what he coins the ‘logic of elimination’ (2006). This logic explains the operation of colonizers arriving and settling on lands already inhabited by another group, specifically that settler colonialism requires the implementation of violence to achieve the goals of a settler society and replace Indigenous structures with those of the settler society. The replacement of Indigenous structures ultimately included the removal and dispossession of Indigenous people themself, which was justified through the racialization of Indigenous people @gonzales_colonialism_2023. Additionally, since the structures of racialization and settler-colonialism inform each other, an analysis of racial decision-making would be incomplete without considering how Indigenous racial formation came to be in the US.

For Indigenous Americans, membership within a tribe is often a key part of this process to distinguish between those who are Indigenous and those who are not. This concept is demonstrated in research done to identify what barriers to tribal enrollment exist for individuals who have been separated from their family and community via adoption and the foster system @landers2021. It was identified that tribal enrollment was key for these individuals to establish and prove their own identity and seek collective identity within the tribe. Essentially, tribal enrollment exists as a “process of mutual verification between the individual and the tribe” (p. 1374).  Additionally, the forceful removal of Indigenous children from their families between the 1800s and 1970s by the United States federal government–-via relocation, boarding schools, child welfare removal, adoption era practices, and so on, has created a group of Indigenous Americans, and their descendants, who have been removed from their family and culture thus further causing turmoil in how they conceptualize their identity. The Federal Acknowledgement Project of 1978, furthered this process since it created an administrative process that required any group wishing to be acknowledged by the federal government as a legitimate tribe to meet certain criteria established by the federal government, further establishing Indigenous American identity as an administrative process rather than of racial, ethnic, or cultural identity @gonzales_eugenics_2007.

Blood quantum often exists as a requirement for tribal membership through a blood quantum minimum which functions as a benchmark of authenticity for who is recognized as Native American by the U.S. government and continues to endure as a key feature of Indigenous identity [@mckay_real_2021 @rodriguez-lonebear_blood_2021]. This system was established as criteria for federal recognition in response to laws like Virginia’s Racial Integrity Act of 1924 and the Indian Reorganization Act (IRA) of 1934 which helped establish racial divisions and criteria on how to quantify Native American identity, respectively @gonzales_eugenics_2007. This establishes that determining Native American identity functions as an administrative process rather than a cultural process of a distinct racial and ethnic group, allowing a governing body to determine who can self-identify a certain way @lambert2019.

-   his would be considered important to talk about engaging with for example when thinking about homelands, ie looking at connections to land

-   should look at churning race and think about how part of the goal of settler colonialism is the replacement of Indigenous people by settlers to become indigenous in their place as a potential explanation? Especially when thinking about hypodescent and why this decision-making process might be occurringIt is key to note that findings made in this paper are not intended to dispute but rather support the findings found in other projects both qualitative and quantitative and base them on the lived reality these studies reflect

    Furthermore, the interpretation of results through the lens of CRT, QuantCrit, or settler-colonialism may not be able to account for every finding however engaging with the social reality Indigenous Americans navigate only serves to expand and explain current understandings regarding Indigenous identity, particularly multi-racial Indigenous identity

    This call for more critically engaged quantitative research is at the basis of this article, meaning that the results will have been interpreted within the context of CRT as well as settler-colonialism due to its focus on Indigenous people. Methods and data

## Sample Selection and Data

The data used for this project was compiled from the American Community Survey data for the years 2010 - 2020. I seek to analyze the race reporting of individuals who indicate Indigenous ancestry in addition to White, Black, or Asian ancestry. For this study, the term Indigenous is used in place of the American Indian and Alaskan Native (AIAN) terms as it is generally more accepted and has grown in popularity in current literature when discussing identity. The sample did not include individuals who listed their race as Latino or Pacific Islander.Part of the data organization of this project required the identification of a multiracial population, specifically an Indigenous multiracial population. To accomplish this goal the Goldstein and Morning method was adopted, which relies on looking at adults' self-reported ancestry, which in turn does not rely on the question of racial identity and instead allows for the construction of a multiracial population based on respondents' self-reported racial ancestry. In regards to the sample, this means looking at respondents who listed Indigenous ancestry in addition to a racially distinct racial ancestry other than Indigenous.Once this group was identified further case selection was performed as these results were compared to their race response. When looking at their racial ancestry and race response, respondents were divided into 4 distinct categories:1) Indigenous alone, which contained respondents that listed their race solely as Indigenous 2) Consistent which included respondents that acknowledged their racial ancestry in full by selecting a racial option that reflected that ancestry 3) Non-Indigenous Race alone, where respondents did acknowledge their Indigenous ancestry when selecting their race response and 4) Inconsistent, where respondents race response did not reflect their racial ancestry in any capacity

Outcome variable: race response Predictor variables: metro, homeland, education above high school, gender and hispanicity These ancestry responses were compared to what their race response and divided into the following groups: Indigenous alone, Multiracial, and Other Race alone where Indigenous alone consisted of those respondents who identified a mixed-ancestry of Indigenous plus one other racial ancestry yet listed their race response as Indigenous alone. Multiracial was those respondents who identified their mixed--Indigenous ancestry and listed a multiracial race response. Other race alone consisted of those respondents with a mixed-Indigenous racial ancestry but listed their racial response as the other race (ie. ancestry listed as Indigenous and Asian but list race as Asian only).

## Methods

The analysis was conducted through creation of several multinomial logit models, which predict the likelihood of a respondent to identify as either Indigenous alone, Multiracial, or other race alone. Three sets of models were created, each set had a different reference category: "Indigenous/White", "Indigenous/Black", and "Indigenous/Asian"

```{r}
#| label: fig-race-ancestry
#| fig-cap: Comparing listed ancestry and race classification
ggplot(data = DATA, aes(x = race_response, fill = race_ancestry, group=race_ancestry, y=after_stat(prop))) + 
  geom_bar(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Self-Identified Racial Classification", y = NULL, fill = "Ancestry")+
  theme(axis.title.x = element_text(vjust = -2))
```

```{r}
```

# Results

# Conclusion
